---
title: "p8105_hw2_dm3175"
author: "Devon Morgan"
date: "9/26/2018"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
```

# Problem 1: NYC Transit Data
Problem 1 will use NYC Transit Data which includes information on entrance and exits for all subway stations located in New York City. 

## Import and Clean Dataset

The dataset was first imported and cleaned. The following variables were retained: line, station name, station latitude/longitude, routes served, entry, vending, entrance type, and ADA compliance. 

```{r import nyc transit data}
NYC_traffic_data = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names() %>% 
select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) %>% 
mutate(entrance_type = tolower(entrance_type),
       line = tolower(line),
       station_name = tolower(station_name))
```

## Description of Dataset



## Reformat Route Number and Route Name Variables 




